# Data visualisation with ggplot2 {#CHAPTER-GGPLOT2}

For this part we use the `FalseBeginners` dataset. The dataset is based on De Wilde et al. [-@dewilde2019], which is available via: <https://osf.io/ndr47/>. I have selected and renamed the variables for our purposes here. @tbl-falsebeginnerscodebook provides an overview of the variables.

| ID  |    Variable    | Description                                                               |
|:-------------:|:-------------:|:------------------------------------------|
|  1  |    Student     | A unique identifier for each student                                      |
|  2  |     School     | A unique identifier for each school                                       |
|  3  |     Class      | A unique identifier for each class                                        |
|  4  |      PPVT      | Score for the Peabody Picture Vocabulary Test (PPVT). Maximum score = 120 |
|  5  |    Speaking    | Score for the speaking test. Maximum score = 20                           |
|  6  |   Listening    | Score for the listening test. Maximum score = 25                          |
|  7  | ReadingWriting | Score for the reading and writing tests. Maximum score = 50               |
|  8  |    Attitude    | Attitude of student towards English: Positive vs. Negative                |
|  9  |       L1       | L1: Dutch vs. Multilingual                                                |
| 10  |      Sex       | Sex of student: Male vs. Female                                           |

: Codebook Falsebeginners {#tbl-falsebeginnerscodebook}

The file is a tab delimited .txt-file. We open the file with `read.delim()`. The dataset is organised in long data format.

```{r}
#| warning: false
library(tidyverse)
```

```{r}
fb <- read.delim("FalseBeginners.txt",                # <1>
                 header = TRUE,                       # <2>
                 na.strings = NA,                     # <3>
                 stringsAsFactors=TRUE)               # <4>
```

1.  the text file to open,
2.  the first row of the file contains the variables names,
3.  missing values are marked as "NA",
4.  strings are interpreted as factors.

As always, we start with a global summary.

```{r}
options(width = 70) 
summary(fb)
```

We omit all rows with NAs (this means that a row is dropped as soon as it contains one missing value).

```{r}
fb <- na.omit(fb)
```

## Histogram

@fig-H4-hist1 shows a histogram for the vocabulary score `PPVT`.

```{r}
#| fig-cap: "A histogram of PPVT"
#| label: fig-H4-hist1
#| fig-width: 3.5
#| fig-height: 2.5
ggplot(data = fb,                 # <2>  
       mapping =  aes(x=PPVT)) +  # <1>
       geom_histogram(bins = 25)  # <3>
```

1.  the dataset that we wish to use. Remember that you need to open this data first. In other words, if the dataframe `fb` does not exist, this code won't work,
2.  mapping & `aes` indicates the variables for the x- and y-axis (and perhaps other variables for other aesthetics),
3.  geom is the kind of visualisation that we wish to use.

The key idea in all ggplot figures is to start from a dataframe or tibble object, choose the variables for the x- and y-axis, and choose a geom, the kind of figure for the visualisation.

Extending the basic code with aesthetic features is quite straightforward. In @fig-H4-hist2, we add some color, a label for the x-axis and use a minimalistic theme.

```{r}
#| fig-cap: "A histogram for PPVT with some additional aesthetics."
#| label: fig-H4-hist2
#| fig-width: 3.5
#| fig-height: 2.5
ggplot(fb,                                  # <1>
       aes(x=PPVT)) +                       # <2> 
  geom_histogram(bins = 25,                 # <3>
                 fill = "#F1A42B") +        # <4>
  theme_minimal() +                         # <5>
  xlab("Score PPVT (120)")                  # <6>
```

1.  the data,
2.  the variable that we wish to visualise,
3.  the visualisation method,
4.  with a colour,
5.  a general theme (there are many others available to accomodate everyone's taste),
6.  a better label for the x-axis.

## Boxplot

The code for the boxplot in @fig-H4-boxplot is very similar to that of the histogram. The only things that we have to change is the use of the y-axis rather than the x-axis and the geom.

```{r}
#| fig-cap: "A vertical boxplot of PPVT"
#| label: fig-H4-boxplot
#| fig-width: 3
#| fig-height: 3

ggplot(fb,                                 # <1>
       aes(y=PPVT)) +                      # <2>
  geom_boxplot() +                         # <3>
  scale_x_discrete( ) +                    # <4>
  theme_grey()                             # <5>  
```

1.  the data,
2.  the variable for the y-axis (one could also create a horizontal boxplot on the x-axis),
3.  use a boxplot,
4.  change the x-axis for a nicer appearance,
5.  use a grey theme.

## Density curve

We change change the geom and we get the density plot in @fig-H4-dens.

```{r}
#| fig-cap: "A density curve of PPVT"
#| label: fig-H4-dens
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb, aes(x = PPVT)) +  # <1>
  geom_density()  +          # <2>
  theme_minimal()
```

## Barplot

Before we can create a barplot we need to tabulate the data, which we can do with `count()`. Then we select the categorical variable levels on the x-axis and their respective counts on the y-axis, and we get @fig-H4-bar1.

```{r}
#| fig-cap: "A barplot of Sex."
#| label: fig-H4-bar1
#| fig-width: 3.5
#| fig-height: 2.5
fb |>                           # <1>
  count(Sex)  |>                # <2>
ggplot(aes(x = Sex,             # <3>
           y = n)) +            # <4>
  geom_col()                    # <5> 
```

1.  start with `fb`,
2.  count the observations for each level of `Sex`,
3.  use ggplot, put `Sex` on the x-axis and `n` (the result of `count()`) on the y-axis,
4.  visualise with a barplot ("col" = column).

We can again add some aesthetics to get @fig-H4-bar2

```{r}
#| fig-cap: "A barplot of Sex with some additional features."
#| label: fig-H4-bar2
#| fig-width: 3.5
#| fig-height: 2.5
fb |>                           
  count(Sex)  |>                
ggplot(aes(x = Sex,             
           y = n)) +            
  geom_col(fill = "#1E64C8") +  
  xlab("Sex")  +           
  theme_minimal()                
```

## Overlapping histograms

So far, we have only looked at univariate visualisations. Let's add an extra dimension. We add `Sex` as a categorical variable by means of two different colours, as in @fig-H4-twohist. The alpha argument argument adds transparency to the colours. Note that we use "fill" rather than "colour", because the latter adds colour to the border of the histogram.

```{r}
#| fig-cap: "PPVT by Sex."
#| label: fig-H4-twohist
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb,                                    # <1>
       aes(x=PPVT,                            # <2>
           fill = Sex)) +                     # <3>
  geom_histogram(bins = 20,                   # <4>
                 position = "identity",       # <5>
                 alpha = 0.4) +               # <6>
  theme_classic()                             # <7>
```

## Overlapping density plots

@fig-H4-twodens shows two overlapping density plots. The code is similar except to the one for the overlapping histograms except for the geom.

```{r}
#| fig-cap: "PPVT by Sex."
#| label: fig-H4-twodens
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb,                                    
       aes(x=PPVT,                            
           fill = Sex)) +                     
  geom_density(position = "identity",       
                 alpha = 0.4) +             
  theme_classic()                           
```

## Multiple boxplots

A clustered boxplot, as in @fig-H4-clusterbox, is created by adding an extra categorical variable on the x-axis. Here we added some extra colour with the `fill` argument.

```{r}
#| fig-cap: "PPVT by Sex."
#| label: fig-H4-clusterbox
#| fig-width: 3.5
#| fig-height: 2.5
ggplot(fb, aes(x = Sex,           
               y = PPVT,            
               fill = Sex)) +         
  geom_boxplot() +                
  theme_minimal() +                
  theme(legend.position = "none")   # <1>
```

1.  A legend is created by default for the colours, but this is redundant because of the labels on the x-axis.

## Boxplots with a facet

@fig-H4-boxfacet visualises three variables: the continuous variable `PPVT` by both `Sex` and `L1`. `L1` is here added as a facet.

```{r}
#| fig-cap: "PPVT by Sex and L1."
#| label: fig-H4-boxfacet
#| fig-width: 3.5
#| fig-height: 2.5
ggplot(fb, aes(x = Sex,           
               y = PPVT,            
               fill = Sex)) +         
  geom_boxplot() +                
  theme_minimal() +   
  facet_wrap(L1 ~ .) +            
  theme_minimal() + 
  theme(legend.position = "none") 
```

@fig-H4-boxfacet2 replaces `L1` by `School`, which conveniently allows us to compare the Sex-differences over the different schools.

```{r}
#| fig-cap: "PPVT by Sex and School."
#| label: fig-H4-boxfacet2
#| fig-width: 12
#| fig-height: 12
ggplot(fb, aes(x = Sex,           
               y = PPVT,            
               fill = Sex)) +         
  geom_boxplot() +                
  theme_minimal() +   
  facet_wrap( ~ School) +            
  theme_minimal() + 
  theme(legend.position = "none") 
```

The plots suggest that in most schools boys tend to have a higher PPVT than girls.

## Clustered barplot

Is there an association between `Sex` and `Attitude`? We can examine the association between two categorical variables by means of clustered barplots, for which we first need to aggregate the data.

```{r}
sa <- fb |> count(Sex, Attitude) 
sa
```

Then we can use the dataframe to create the barplot in @fig-H4-clustbar.

```{r}
#| fig-cap: "A clustered barplot of Sex and Attitude"
#| label: fig-H4-clustbar
#| fig-width: 4
#| fig-height: 2
ggplot(data=sa, aes(x=Sex,             
                    y=n,               
                    fill=Attitude)) +  
  geom_col(position = "dodge") +       
  theme_minimal()
```

We can aggregate the attitudes for both Sexes by School and add a facet for `School` to the clustered barplot, as in @fig-H4-clustbarfacet

```{r}
#| fig-cap: "Attitude by Sex for each School"
#| label: fig-H4-clustbarfacet
#| fig-width: 12
#| fig-height: 12
ssa <- fb |> group_by(School, Sex) |> count(Attitude)
ggplot(data=ssa, aes(x=Sex,             
                    y=n,                
                    fill=Attitude)) +   
  geom_col(position = "dodge") +        
  facet_wrap( ~ School) + 
  theme_minimal()
```

Overall, both sexes show a very positive attitude towards English.

## Scatterplot

The scatterplot in @fig-H4-scatter visualises the relation between the scores for `PPVT` and `Speaking`.

```{r}
#| fig-cap: "A scatterplot of PPVT and Speaking."
#| label: fig-H4-scatter
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb, aes(x = PPVT,          
               y = Speaking)) +   
  geom_point()   +                
  theme_minimal()
```

We can add a regressionline with `geom_smooth()` as in @fig-H4-scatterline.

```{r}
#| fig-cap: "A scatterplot of PPVT and Speaking with a regression line"
#| label: fig-H4-scatterline
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb, aes(x = PPVT,          
               y = Speaking)) +   
  geom_point()   +              
  geom_smooth(method = "lm", col="blue")  +
  theme_minimal()
```

Or with a non-parametric smoother, as in @fig-H4-smoother.

```{r}
#| fig-cap: "A scatterplot of PPVT and Speaking with a smoother."
#| label: fig-H4-smoother
#| fig-width: 3.5
#| fig-height: 2.5

ggplot(fb, aes(x = PPVT,          
               y = Speaking)) +   
  geom_point()   +        
  geom_smooth(method = "loess", col="blue")  +
  theme_minimal()
```

And again, it's straightforward to add extra variables a facet, as in @fig-H4-scatterschool:

```{r}
#| fig-cap: "A scatterplot of PPVT and Speaking, by Sex and L1."
#| label: fig-H4-scatterschool
#| fig-width: 6
#| fig-height: 6

ggplot(fb, aes(x = PPVT,          
               y = Speaking)) +   
  geom_point()   +  
  geom_smooth(method = "loess", col="blue")  +
  facet_grid(Sex ~ L1) + 
  theme_minimal()
```

## Exercises

### **Exercise 4.1** {.unnumbered}

Use the ggplot2 package to create the following visualisations for the `Multitask` dataset:

1.  Create a histogram of `Time`
2.  Create a boxplot of `Time`
3.  Create a barplot of `Location`

### **Exercise 4.2** {.unnumbered}

Use the ggplot2 package to create the following visualisations for the `Multitask` dataset:

1.  Visualise `Time` by `Task` as a clustered boxplot
2.  Visualise `Time` by `Task` and University as overlapping density plots
3.  Add a vertical line to visualize the mean `Time` for both Tasks.
4.  Use `ggplot()` to visualise the paired differences in the multitask dataset
5.  Create a new variable `Difference`, based on the difference in `Time` between the two Tasks.
6.  Visualise the relation between `Speaking` and `Listening`.
7.  Visualise the relation between `Speaking` and `Listening` for both Sexes and L1s separately.
